Automated Lensing Learner - I: An automated Strong Lensing Identification Pipeline

### 2017-03-14 ms.tex
Marked the to-do for final intro paragraph to resolve
Streamlined the mock observations description in Sect 2.1, also broke it into subsubsections for readability

% Convolutional and residual
In image space, \citep{petrillo_etal17} used data from the Kilo Degree
survey to train convolutional neural networks with the Kilo Degree
Survey. \citet{bom_etal16}. \citet{lanusse_etal17} using
state-of-the-art residual networks to....  directly in the image space
with minimal image pre-processing.

%  Relevance of our work
While there has been a recent surge in the use of neural networks
applied to image classification problems, \wording{these
  implementations are not readily scalable, nor are the necessary
  computational resources and expertise easily accessible to the
  entire scientific community.}  We present a supervised
classification pipeline built with open source tools, tested on mock
HST and LSST data, which will also be made publicly available.  Our
goal is to centralize an open source pipeline that is general enough
for the user to select or add appropriate image processing techniques
to augment the features of the lensed image, train and test data with
different machine learning techniques, and to select the sequence of
methods that provides optimal model performance.  Results are
reproducible on personal computers, as both the pipeline and the data
will be publicly distributed.

@@

%-------------------------------------------------%
\section{Methodology}
\label{sec:methods}
%-------------------------------------------------%
\subsection{Creating Simulated Images for Training Set and Testing Set}

We use mock Hubble Space Telescope (HST) and Large Synoptic Sky
Telescope (LSST) images generated with PICS (Pipeline for Images of
Cosmological Strong lensing) \citep{li_etal16} and LensPOP
\citep{collett_15} as training set and testing set. In these images,
the lensed images include the images of lens galaxies, lensed images,
and galaxies on the line of sight; the unlensed images include the
images of lens galaxies and the galaxies on the line of sight.

To produce simulated lensed galaxies, we first set mass models of the
lenses to be the Singular Isothermal Ellipsoid (SIE):

\begin{equation}
\kappa = \frac{\theta_{\rm E}}{2}\frac{1}{\sqrt{x_1^2/q+x_2^2 q}}.
\end{equation}

Where, $\theta_{\rm E}$ is the Einstein Radius, $q$ is axis
ratio. Einstein radius can be calculated according to the redshift of
the lens, the redshift of the source, and the velocity dispersion of
the lens galaxy.
\begin{equation}
\theta_{\rm E} = 4\pi\left(\frac{\sigma_v}{c}\right)\frac{D_{ls}}{D_{s}}, 
\end{equation}
where, $c$ is the speed of light, $\sigma_v$ is the velocity
dispersion of the lens galaxy, $D_{ls}$ and $D_{s}$ are the angular
diameter distance from the source plane to the lens plane and from the
source plane to the observer separately.

To rotate the lenses with a random orientation angles, we adopt the
transformation below:

\begin{eqnarray}
\label{eq:rotation}
\begin{bmatrix}

    x_{1}\prime      \\
    x_{2}\prime 
\end{bmatrix}
= 
\begin{bmatrix}
    \cos \phi  &  -\sin \phi      \\
    \sin \phi  &  ~\cos \phi      
\end{bmatrix} 
\begin{bmatrix}
    x_1      \\
    x_2      
\end{bmatrix} .
\end{eqnarray}

$\phi$ is the orientation angle.  According to above equations, we
need $\sigma_v$, $q$, $\phi$, redshift of the lens $z_l$, and redshift
of the source $z_s$ to derivative the lensing maps completely.
Velocity dispersion, ellipticity, and orientation are randomly chosen
from typical ranges of a galaxy, where $\sigma_v = [200, 320] km/s$,
$q = [0.5, 1.0]$, and $\phi = [0, 360]$. The redshift of the lens is
fixed at $z_l = 0.2$, and the redshift of the source is fixed at $z_s
= 1.0$. The images of source galaxies are extracted from the CANDELS
survey. We selected 33 bright galaxies, and relocated them to
projected positions, which are close to the caustic of the lensing
system, and the projected position of the lens is fixed in the center
of the field of view of each image. Once we have the lens, source, and
the parameters of the image simulation, i.e., $300 \times 300 {\rm
  pixels}^2$ and 0.03 arcsec per pixel per side, settled, the lensed
images can be produced by performing ray-tracing simulations.

The light distributions of the lens galaxies  are modeled as the elliptical Sersic profile,

\begin{equation}
I(R) = I_{\rm eff}~{\rm exp} \left\lbrace -b_{n} \left[ \left(
  \frac{R}{R_{\rm eff}}\right)^{1/n} - 1 \right ] \right\rbrace
\end{equation}

Where, $R = \sqrt[]{x_1^2 /q+x_2^2 q }$, $R_{\rm eff}$ is the
effective radius in units of arcsecond, $I_{\rm eff}$ is the intensity
at the effective radius, $n$ is the index of the Sersic profile, $q$
is the ellipticity, and We perform the similar transformation as
Eq. \ref{eq:rotation} to orient the sources.  We assume that the
distribution of light is following that of mass, so ellipticity and
orientation are the same as the SIE model. Matching the velocity
dispersion with the COSMOS morphological
catalog \footnote{\url{https://arxiv.org/pdf/1501.04977v2.pdf}}, we
assign effective radius, effective luminosity, and index to the light
profile.  We also assume the light center is on top of the mass
center, the noiseless images of lenses can be drawn. The galaxies on
the line of sight are constructed by cutting light-cones from the
Hubble Ultra Deep Field.  Stacking all these images together and
collaborating the magnitude of each component, we create primarily
simulated lensed images.

The final part is mock observing. To produce HST-like images, we do
not need to do anything further than stacking and magnitude
calibrating, because noise and PSF are involved when we are extracting
galaxies on the line of sight. To produce LSST-like images, we perform
re-binning, convolving with PSF, and adding noise serially; this
process is implemented in LensPop. We created single-epoch-LSST-like
images and stacked -LSST-like images for investigating the performance
and usability of our pipeline in LSST data. Figure 2.1 illustrates a
sample mock lensed and unlensed images from each telescope. From left
to right, the top panel images respectively correspond to a mock HST
lensing and non-lensing system. The mock HST images are 300 × 300 in
size, and the pixel size is 0.03 arcsecond. For the HST-like data-set,
many giant arcs are visually obvious. The middle panel images
respectively correspond to a mock single-epoch LSST lensing and
non-lensing systems. The bottom panel images respectively correspond
to a mock stacked LSST lensing and non-lensing systems. These images
are 45 × 45; the pixel size is 0.2 arcsecond. The ground based noise,
PSF, and limited resolution of the LSST-like data-set make visual
giant arc identification difficult. But when in the stacked LSST
images, the quality is better, because more data improves the signal
to noise ratio better.

In the training sets, we also generate unlensed images, the procedure
is similar to mock lensed images but getting rid of the part of
ray-tracing. 10000 lensed images and 10000 unlensed images are merged
to construct one training set. Furthermore, for investigating the
influence of BCG galaxies on the performance of our lens
identification pipeline in LSST data, we build another training sets
of the images without BCG galaxies.

%-------------------------------------------------%
\subsection{Mock Images}
%-------------------------------------------------%

% Summary of HST and LSST images
We use mock {\em Hubble Space Telescope} (HST) and {\em Large Synoptic
  Sky Telescope} (LSST) images generated with PICS (Pipeline for
Images of Cosmological Strong lensing) \citep{li_etal16} and
LensPop\citep{collett_15}.  These images include cluster-member
galaxies and foreground stars.  Half of the images used are strong
lensing systems.  Here, lensed galaxy images are ray-traced images of
actual galaxies extracted from deep Hubble Space Telescope
observations.  The final mock images are ``observed'' with a realistic
point spread function corresponding to modeled detector artifacts for
bright stars.  See \cite{li_etal16} for more details.

Figure ~\ref{fig:mockimages} illustrates a sample mock lensed and
unlensed images from each telescope.  The top left and right panel
images respectively correspond to a mock HST lensing and non-lensing
system. The mock HST images are $n_\text{pix}\times
n_\text{pix}=300\times300$ in size.  For the HST-like dataset, many
giant arcs are visually obvious.

The middle row corresponds to the same simulated systems, as observed
with a single epoch of LSST.  These images are $n_\text{pix}\times
n_\text{pix}=45\times45$.  The ground based noise and limited
resolution of the LSST-like dataset make visual giant arc
identification difficult.  However, by stacking observations over the
course of the LSST survey, we recover the arc feature, albeit at a
much lower resolution than with the HST-like image. The bottom row
shows the mock stacked observation, where the lensed background galaxy
is now visible.

@@

%-------------------------------------------------%
\subsection{Strong Arc Lensing Identification Pipeline}
%-------------------------------------------------%

>>>>

% Convolutional and residual
In image space, \citep{petrillo_etal17} used data from the Kilo Degree
survey to train convolutional neural networks with the Kilo Degree
Survey. \citet{lanusse_etal17} using state-of-the-art residual
networks to work directly in the image space with minimal image
pre-processing.  A major strength of the \citet{lanusse_etal17}
implementation is that residual networks ..... , which regular
convolutional networks do not.

EDITING HERE: Double check that petrillo paper
operates in image space.

%  Relevance of our work
While there has been a recent surge in the use of neural networks
applied to astronomy image classification problems, \wording{these
  implementations are not readily scalable, nor are the necessary
  computational resources and expertise easily accessible to the
  entire scientific community.}  We present a supervised
classification pipeline built with open source tools, tested on mock
HST and LSST data, which will also be made publicly available.  Our
goal is to centralize an open source pipeline that is general enough
for the user to select or add appropriate image processing techniques
to augment the features of the lensed image, train and test data with
different machine learning techniques, and to select the sequence of
methods that provides optimal model performance.  Results are
reproducible on personal computers, as both the pipeline and the data
will be publicly distributed.

@@

%-------------------------------------------------%
\section{Methodology}
\label{sec:methods}
%-------------------------------------------------%
\subsection{Mock Observations}
%-------------------------------------------------%
\subsection{Mock Images}
%-------------------------------------------------%
% Summary of HST and LSST images
We use mock {\em Hubble Space Telescope}\footnote{\todo{hst...org}}
(HST) and {\em Large Synoptic Sky
  Telescope}\footnote{\todo{lsst....org}} (LSST) images generated with
PICS (Pipeline for Images of Cosmological Strong lensing)
\citep{li_etal16} and LensPOP \citep{collett_15} to train and test our
model.  Along the line of sight galaxies and foreground stars.  Half
of the images used are strong lensing systems.  Lensed galaxy images
are ray-traced images of actual galaxies extracted from deep Hubble
Space Telescope observations.  The final mock images are ``observed''
with a realistic point spread function corresponding to modeled
detector artifacts for bright stars.  We refer to \cite{li_etal16} for
further details, but outline the process here.

We generate 10,000 lens containing mock observations and 10,000
non-lens containing mock observations to run our parameter search, and
keep a hold-out test set of 1,000 lens containing mock observations
and 1,000 non-lens containing mock observations on which we evaluate
the final trained model.

Lens containing mock observations include the lens galaxy, lensed
images of the source galaxy, and galaxies along the line of sight.
Non-lens containing mock observations include all but a lensed source
image.  We convolve each of the $2\times 10,000$ train/test images and
$2\times 1,000$ hold-out images to produce three separate sets of mock
``observations'' with equal numbers of lens and non-lens containing
systems: (1) HST-like observations, (2) LSST-like observations over
the span of one year, and (3) LSST-like observations over the span of
ten years.  We respectively label these observations as HST, LSST1,
and LSST10.
%-------------------------------------------------%
\subsubsection{Modeling Images of the Source Galaxies}
%-------------------------------------------------%
To produce images of the simulated lensed source galaxies, we first
set mass models of the lens galaxy to be the Singular Isothermal
Ellipsoid (SIE):
%-------------------------------------------------%
\begin{equation}\label{eqn:sie}
\kappa = \frac{\theta_{\rm E}}{2}\frac{1}{\sqrt{x_1^2/q+x_2^2 q}},
\end{equation}
%-------------------------------------------------%
where $\theta_{\rm E}$ is the Einstein radius and $q$ is the axis
\wording{ratio between semi-major and semi-minor axes $x_1$ and
  $x_2$}. We can calculate the Einstein radius from the redshift of
the lens, the redshift of the source, and the velocity dispersion of
the lens galaxy.
%-------------------------------------------------%
\begin{equation}\label{eqn:einsteinradius}
\theta_{\rm E} = 4\pi\left(\frac{\sigma_v}{c}\right)\frac{D_{ls}}{D_{s}}. 
\end{equation}
%-------------------------------------------------%
Here, $c$ is the speed of light, $\sigma_v$ is the velocity dispersion
of the lens galaxy, and $D_{ls}$ and $D_{s}$ are the respective
angular diameter distance from the source plane to the lens plane and
from the source plane to the observer.

To rotate the lenses with a random orientation angles, we adopt the
transformation below for orientation angle, $\phi$:
%-------------------------------------------------%
\begin{eqnarray}
\label{eqn:rotation}
\begin{bmatrix}

    x_{1}\prime      \\
    x_{2}\prime 
\end{bmatrix}
= 
\begin{bmatrix}
    \cos \phi  &  -\sin \phi      \\
    \sin \phi  &  ~\cos \phi      
\end{bmatrix} 
\begin{bmatrix}
    x_1      \\
    x_2      
\end{bmatrix} .
\end{eqnarray}
%-------------------------------------------------%
From Equations~\ref{eqn:sie}~-~\ref{eqn:rotation}, there are five
independent parameters that uniquely define the lensing system:
$\sigma_v$, ellipticity $q$, orientation angle $\phi$, redshift of the
lens $z_l$, and redshift of the source $z_s$.

% Parameters of lensing system
We choose $\sigma_v$, $q$, and $\phi$ randomly from typical ranges of
a galaxy, where $\sigma_v = [200, 320] km/s$, $q = [0.5, 1.0]$, and
$\phi = [0, 360]$.  The redshift of the lens galaxy is drawn from the
distribution \todo{found in \citet{???}}.  We fix the redshift of the
source to $z_s = 1.0$. We then extract images of 33 bright galaxies
from the CANDELS survey as source galaxies.  We place the source
galaxy images in projected positions close to the caustic of the
lensing system and fix the projected position of the lens galaxy to
the center of the field of view of each image.  Once we have
determined all parameters of the lensing system, we perform
ray-tracing simulations to produce the final lensed image.

%-------------------------------------------------%
\subsubsection{Modeling Lens Galaxies}
%-------------------------------------------------%
We model the light distributions of the lens galaxies with an
elliptical Sersic profile,
%-------------------------------------------------%
\begin{equation}\label{eqn:sersic}
I(R) = I_{\rm eff}~{\rm exp} \left\lbrace -b_{n} \left[ \left(
  \frac{R}{R_{\rm eff}}\right)^{1/n} - 1 \right ] \right\rbrace,
\end{equation}
%-------------------------------------------------%
where, $R = \sqrt[]{x_1^2 /q+x_2^2 q }$, $R_{\rm eff}$ is the
effective radius in units of arcseconds, $I_{\rm eff}$ is the
intensity at the effective radius, $n$ is the index of the Sersic
profile, and $q$ is the ellipticity of the lens galaxy.  We perform a
similar transformation as Equation~\ref{eqn:rotation} to orient the
sources.  We assume that the distribution of light is following that
of mass, so ellipticity and orientation are the same as the SIE model.

We use the COSMOS morphological
catalog \footnote{\url{https://arxiv.org/pdf/1501.04977v2.pdf}} to
match the velocity dispersion with an assigned effective radius,
effective luminosity, and index to the light profile.  We also assume
the light center is on top of the mass center, creating noiseless
images of lens galaxies.

We construct galaxies along the line of sight by cutting light-cones
from the \todo{Hubble Ultra Deep Field\footnote{udf...org}}.  We stack
these images with the lens galaxy image and the lensed source galaxy,
calibrating the magnitude of all three components.

%-------------------------------------------------%
\subsubsection{Producing Mock Images}
%-------------------------------------------------%

The final step is to perform a mock observations.  For HST-like
observations, galaxies along the line of sight already include noise
and a point spread function (PSF).  We therefore stack and magnitude
calibrate all components to produce the final HST-like images, which
are $300\times300~{\rm pix}^2$ with $0.03$~arcsec per pixel.

To produce LSST-like images, we use {\em LensPop} to serially perform
re-binning, convolve with the corresponding PSF, and add
noise\citep{collett_15}.  We created one year LSST-like (LSST1) images
and stacked ten-year LSST-like images (LSST10) to investigate the
performance and usability of our pipeline in future LSST
data. LSST-like images are $45\times45~{\rm pix}^2$ with $0.2$~arcsec
per pixel.

Figure~\ref{fig:mockimages} illustrates sample mock observations with
a strong lensing signature from each telescope.  The left-most column
respectively correspond to a mock HST lensing system with a highly
magnified source galaxy (top) and a less visible image of the source
galaxy (bottom). For the HST-like dataset, many arcs are visually
obvious due to the exquisite spatial resolution.

The middle column corresponds to the same simulated systems, as
observed with over ten years of LSST, LSST10.  The right-most column
corresponds to the simulated systems observed over just one year of
LSST1.  LSST10 images visually exhibit an improved signal to noise
ratio, recovering the arc feature, albeit at a much lower resolution
than with the HST-like image.  The top images of LSST10 and LSST1 show
a visible lensed source galaxy image.  The bottom images are washed
out in the bottom row, where the magnification of the source galaxy is
not as large.  The ground based noise, PSF, and limited resolution of
the LSST1 make visual giant arc identification difficult, except in
systems with the most magnified source galaxies.

@@

Our mock observations also include non-lens containing images.  The
procedure is similar to mock lensed images but we do not perform
ray-tracing, so these images do not have lensed source galaxies.

Furthermore, we investigate the influence of light from the lens
galaxies on the performance of our lens identification pipeline.  We
generate another set of each HST, LSST1, and LSST10 images without the
lens galaxy.  We label these nHST, nLSST1, and nLSST10.

The final data is then comprised of $6\times10,000$ lens containing
images, and $6\times 10,000$ non-lens containing images.  We also keep
a hold-out set of $6\times 1,000$ lens containing images, and $6\times
1,000$ non-lens containing images.
%-------------------------------------------------%
\subsection{Strong Arc Lensing Identification Pipeline}
%-------------------------------------------------%
